---
title: "Exploratory-Data-Analysis-R Part 2"
author: "Illarion Jabine"
date: "4/5/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

### Required packages:

* [DataExplorer]
* [GGally]
* [graphics]: base R function pairs()
* [MASS] : dataset Car93
* [tidyverse]

[DataExplorer]
Automated data exploration process for analytic tasks and predictive modeling.
The package scans and analyzes each variable, and visualizes them with typical graphical techniques.
Common data processing methods are also available to treat and format data.

[GGally]
'GGally' extends 'ggplot2' by adding several functions to reduce the complexity of combining geometric objects with transformed data.
Some of these functions include:
 * a pairwise plot matrix, 
 * a two group pairwise plot matrix, 
 * a parallel coordinates plot, 
 * a survival plot, 
 * several functions to plot networks.

```{r}
library(DataExplorer)
library(GGally)
library(MASS)
library(tidyverse)
```

Get Car93 data set from MASS package and then detach it.
```{r}
data(package = "MASS")
car <- Cars93
detach("package:MASS", unload = TRUE)
```

Let's insert some NAs in car data set and save it as car_with_NAs
```{r}
car_with_NAs <- edit(car)
sum(is.na(car_with_NAs))
```

### 1. Data Explorer package

Function create_report() from DataExplorer package creates a data profiling report. A separate html file is generated with the various elements:

•Basic Statistics 
 ◦Raw Counts
 ◦Percentages
•Data Structure
•Missing Data Profile
•Univariate Distribution 
 ◦Histogram
 ◦Bar Chart (with frequency)
 ◦QQ Plot
•Correlation Analysis
•Principal Component Analysis

```{r}
create_report(car_with_NAs)
```

You can also produce individual reports by calling various plot_ functions:
```{r}
# Plot density estimates for each continuous feature
plot_density(car_with_NAs %>% select(Price,RPM,Rev.per.mile))

# Describe basic information for input data
plot_intro(car_with_NAs)
```

### 2. GGally package

This ggplot2 extension package has numerous ggally_ helper functions, that run on top of ggplot2.
I personally find ggpairs() ggplot2 generalized pairs plot quite powerfull.
This function is similar to standard R matrix of scatterplots produced by pairs(), but shows extra information like density plots, correlation coefficient, etc:

```{r}
pairs(x = subset(car, select = c(MPG.city,Price,Horsepower)))
ggpairs(data = subset(car, select = c(MPG.city,Price,Horsepower)))
```
ggpairs() is quite intelegent though, if I add a factor variable "Cylinders" it will understand this fact and adds a bar plot rather than a density plot and box plots in erspective cells:  
```{r}
ggpairs(data = subset(car, select = c(MPG.city,Price,Horsepower,Cylinders)))
```

