---
title: "Exploratory Data Analysis in R using dlookr package"
author: "Illarion Jabine"
date: "10/2/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

### Required packages:

* [tidyverse]
* [dlookr]: Tools for Data Diagnosis, Exploration, Transformation
```{r}
vignette(package="dlookr")
```

* [flextable]: Functions for Tabular Reporting. Makes an output prettier.
* [missForest]: prodNA() function

### Useful links
https://yuzar-blog.netlify.app/posts/2021-01-30-r-package-reviews-dlookr-diagnose-explore-and-transform-your-data/

```{r Loading libraries and data}
library(tidyverse)
library(dlookr)
library(flextable)
library(missForest)
data("diamonds") # dataset from ggplot2
df <- diamonds
rm(diamonds) # remove diamonds dataset
```

### 1. 5 number statistics with dlookr package

describe() function returns descriptive statistics of all numeric variables:
```{r}
describe(df)
```

dlookr also has various plot functions:
plot_bar_category() - Plot bar chart of categorical variables
plot_box_numeric() - Plot Box-Plot of numerical variables
plot_hist_numeric() - Plot histogram of numerical variables
```{r}
plot_hist_numeric(df)

# can we have the same with ggplot? Yes we can!
df.num <- df %>% select_if(is.numeric) 
df.num %>% pivot_longer(colnames(df.num)) %>%
  ggplot(aes(x = value)) +    # Draw each column as histogram
  geom_histogram() + 
  #scale_y_continuous() +
  facet_wrap(~ name, scales = "free")

```


### 2. General data quality check

dlookr package has a set of generic diagnostic functions.
diagnose() - generic data quality information, like missing and unique values:
```{r}
diagnose(df) %>% flextable()
```
diagnose_category() - checks all categorical data in a dataset.
```{r}
diagnose_category(df) %>% flextable()
df %>% select_if(is.factor) %>% group_by(cut) %>% summarise(n=n())
```
diagnose_numeric() - diagnose all numeric variables:
```{r}
diagnose_numeric(df) %>% flextable()
```
Basically we can select only numeric columns (or categorical) using select_if() or summarise_if() or across() from dplyr:
```{r}
df %>% select_if(is.numeric) %>% summarise(across(.cols = everything(),list(min = min, mean = mean, median = median), .names = "{.col}.{.fn}"))
df %>% select_if(is.numeric) %>% summarise_all((list(min,max)))
```


### 3.Outlier analysis

```{r}
df %>% diagnose_numeric() %>% filter(variables == "x") %>% select(outlier) 
```

In x variable there are 32 outliers. We can analyze them further using R base functionality:

```{r}
boxplot.res <- boxplot(df$x)
length(boxplot.res$out) # number of outliers in x variable. We pull from an object created by boxplot() function. Very handy.
# Or we can calculate outliers like that:
df$x[(df$x < quantile(df$x,0.25) - 1.5*IQR(df$x))|(df$x > quantile(df$x,0.75) +1.5*IQR(df$x))]

```

dlookr package also has a funcition to analyze outliers.

To locate what columns contain outliers:
```{r}
find_outliers(df,index = FALSE) # if index = FALSE, function returns column names.
```

diagnose_outlier() as output produces outliers count, outliers ratio, outliers mean, mean with and without outliers.

```{r}
diagnose_outlier(df) %>% flextable()
# better with rounding 
diagnose_outlier(df) %>% mutate(across(where(is.double),round))  %>% flextable()
```
From outlier diagnose we can clearly see that price variable is heavily influenced by them.

Should yo want to plot box plots and histograms with and without outliers, use plot_outlier() function:
```{r}
df %>% plot_outlier() # for all numeric variables
df %>% select(price) %>% plot_outlier() # for price only
```

### 4. Missing values

Let's first check if there are any missing values:
```{r}
sum(is.na(df))
summary(df)
```

No missing values. Let's insert some using prodNA() from missForest package.
prodNA <-
function (x, noNA = 0.1) 
{
    n <- nrow(x)
    p <- ncol(x)
    NAloc <- rep(FALSE, n * p)
    NAloc[sample(n * p, floor(n * p * noNA))] <- TRUE
    x[matrix(NAloc, nrow = n, ncol = p)] <- NA
    return(x)
}
```{r}
df1 <- prodNA(df, noNA = 0.2) # 20% proportion of missing values w.r.t. the number of entries of 'x'.
length(which(is.na(df1)))/(dim(df1)[1]*dim(df1)[2]) # yeap, 0.2
rm(df1)

# create a copy of df
df.na <- df 

# Let's now insert NAs into a new df.na such that in first column there are 50 % NAs, in second column - 40 %, etc. For that we construct a vector of NAs %:
na.percent <- c(.50,.40,.30,.20,.15,.10,.5,0,0,0)

# in for loop assign NAs with % from na.percent vector
for (i in seq_along(df)) {
  df.na[,i] <- prodNA(df[,i],na.percent[i])
}
```
dlookr can produce 3 missing value plots.

1. plot_na_pareto():	Pareto chart for missing value
```{r}
plot_na_pareto(df.na)
plot_na_pareto(df.na,plot = FALSE) # plot = FALSE produce a table as output.
plot_na_pareto(df.na,only_na = TRUE, plot = FALSE) %>% flextable() # only shows columns with NAs
```
Indeed pareto chart shows percentages as per na.percent vector.

2. plot_na_hclust():	Combination chart for missing value
```{r}
plot_na_hclust(df.na)
```
Unfortunately running plot_na_hclust(df.na) with dim(df.na):
[1] 53940    10
Produced an error: "Error: cannot allocate vector of size 9.5 Gb"
Let's remover all objects and recreate those required:
```{r}
rm(list = ls())
df <- diamonds
na.percent <- c(.50,.40,.30,.20,.15,.10,.5,0,0,0)
df.na <- df 
# in for loop assign NAs with % from na.percent vector
for (i in seq_along(df)) {
  df.na[,i] <- prodNA(df[,i],na.percent[i])
}
gc()
memory.limit(size = 30000)
```




plot_na_intersect():	Plot the combination variables that is include missing value