---
title: "Missing Value Analysis in R"
author: "Illarion Jabine"
date: "13/12/2019"
output:
  word_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
	echo = TRUE,
	message = FALSE,
	warning = FALSE
)
```


### Required packages:

* missForest: to generate NAs at random
* mice: for missing value patters
* VIM: Visualization and Imputation of Missing Values
* dplyr: select_if() - to select only numeric variables
* Hmisc: imputing missing values
* caret: kNN imputation

## Missing Value Analysis and Handling

Missing values is a common problem, and might cause serious problem down the road during the model building phase.
This issue is pervasive and difficult to deal with. 
Generally speaking missing data can be in one of the following categories:
 1. Missing Completely at Random (MCAR)
 2. Missing at Random (MAR)
 3. Missing Not at Random (MNAR)
* When we have MCAR there is no way to predict where in the data we can come across the missing data. To handle this situation you just drop rows with the missing values.
* MAR situation is when missingness is associated with other variables. That actually means that conditioned on some variable in the data set the data are missing completely at random.
To deal with MAR we can perform multiple imputation to predict what missing values could be. 
There are many algorithms to perform imputations which allow to capture statistical structure of the relationships between the variables.
* MNAR is when the missing data are neither MCAR nor MAR. It is often caused by deterministic relationship between variables.

So, handling this issue is of paramount importance. First of all you need to identify if a dataset contains any missing values. Once it's done you have 4 ways to handle them:
1. Go back to the source of data (if possible) and find out what causes the missing values.
2. Delete the observations with missing values.
3. Delete the variables with missing values.
4. Replace the missing values with the mean or median.
5. Predict missing values by imputing them using methods like kNN, random forest, tree, lm, etc.


As an example, let's create a dataset with missing values. We will use prodNA() function from lissForest package to generate random NAs:

```{r create dataframe with NAs}
library(missForest)
library(mice)
library(VIM)
library(dplyr)
library(tidyverse)
library(Hmisc)
library(caret)
data(iris)

# create a new dataframe with 10% missing values:
iris.mis <- prodNA(iris, .10)

# another way using sample() function to enject missing values:
iris.mis1 <- iris
iris.mis1[sample(1:nrow(iris), round(nrow(iris)/10)), "Sepal.Length"] <- NA
iris.mis1[sample(1:nrow(iris), round(nrow(iris)/10)), "Sepal.Width"] <- NA
iris.mis1[sample(1:nrow(iris), round(nrow(iris)/10)), "Petal.Length"] <- NA
iris.mis1[sample(1:nrow(iris), round(nrow(iris)/10)), "Petal.Width"] <- NA
```

## 1. Missing Value Initial Analysis

The first step is to analyse if the dataset contains missing values,what variables are affected and the distribution pattern.

The basic functions that allow to detect the presence of missing values incluce summary(), colSums()
```{r detecting missing values, echo=TRUE}
# Quicly check if there exist any NAs using standard anyNA() function from base package:
anyNA(iris.mis)

# shows the columns with missing values and their count.
summary(iris.mis)
# To see the total number of missing values:
sum(is.na(iris.mis))

# we select only numeric colums and checks if they contain NAs
select_if(iris.mis,is.numeric) %>% colSums()

# If we want to see the total count of NAs in each column:

# apply(2, is.na) - returns a logical vector of TRUE and FALSE for each column, and apply(2, sum) sums up (basically counts TRUEs (i.e. NAs)):
total_nas_by_column <- iris.mis %>% apply(2, is.na) %>% apply(2, sum)

# show the bar plot with NAs counts in each variable:
barplot(total_nas_by_column)


# another way to select only numeric columns:
# iris.mis[, unlist(lapply(iris.mis, is.numeric))]

# More advanced way to see the structure of the missing data is to use md.pattern() from mice package:
mice::md.pattern(iris.mis)
# the last row of the matrix contains the number of missing values in each column

# To see missing values in more visual format use aggr() from VIM package:
# This function calculates or plots the amount of missing/imputed values in each variable and the amount of missing/imputed values in certain combinations of variables.

missing_value_analysis <- aggr(iris.mis, col=c("skyblue", "red"),
 numbers=TRUE, sortVars=TRUE,
 labels=names(iris.mis), cex.axis=.7,
 gap=2, ylab=c("Missing Data Histogram","Pattern"))

# To see summary of the missing values analysis produced by aggr use generic function summary():
summary(missing_value_analysis)

```

## 2.Replacing Missing Values with mean or median

Replacing NAs with mean or median values is quite a straightforward, brute force approach. It is not as sophisticated as predicting the missing values by means of prediction methodes, but as a quick and dirty approach it might do. 

```{r Replacing Missing Values with mean or median}
# Let's first exclude the categorical variable Species from our computation:
iris.mis <- iris.mis[,-5] 
# We can exclude Species column using subset function: iris.mis <- subset(iris.mis, select = -c(Species))


# Let's replace NAs with column vector means using mice() function from mice package:
mean_compute_mice <- mice::mice(data = iris.mis, method = "mean")

# To retrive the new dataset with computed NAs from the object mean_compute_mice use complete() function:
mean_compute_mice_res <- mice::complete(mean_compute_mice)

# We can impute NAs using impute() function from Hmisc package. If instead of function a single value is specified those value is used for insertion.

impute(iris.mis$Sepal.Length, mean) # replace NAs with mean
impute(iris.mis$Sepal.Length, median) # replace NAs with median
impute(iris.mis$Sepal.Length, 10) # replace NAs with 10

# Or we can replace missing values with minimal or maximal values:
impute(iris.mis$Sepal.Length, min)
impute(iris.mis$Sepal.Length, max)


# Replacing using R standard functionality:

# iris.mis$Sepal.Length[is.na(iris.mis$Sepal.Length)] <- mean(iris.mis$Sepal.Length, na.rm = TRUE)
```

## 3. Impute Missing Values by using prediction algorithm

This is more sophisticated methode of estimating values for the missing values by applying different prediction models learned for missing values. Quite popular NAs imputational methodes include kNN, classification and regression trees, linear and logistic regression, random forest, just to name a few. 
Note: For available imputation methodes, see help(mice).

Let's compute missing values using different prediction algorithms:

1. NAs replaced with the mean of the k nearest neighbors. 
I have found several packages that can perform kNN imputation (impute, DMwR, caret).
Using caret package takes several steps:

```{r missing value calculation: kNN with caret}
#load caret library
library(caret)
# run preProcess() to compute missing values with the means of k=10 nearest  neighbors.
Knn_NAs_preProcess <- preProcess(iris.mis,
                            method = c("knnImpute"),
                            k = 10,
                            knnSummary = mean)
# preProcess() automatically center and scale the data, regardless of what is in the  method  argument.

# To create a transformed dataset use predict() function with Knn_NAs_preProcess as one of the parameters:
Knn_NAs_calculation <- predict(Knn_NAs_preProcess, iris.mis,na.action = na.pass)

# We need to de-normalize Knn_NAs_calculation dataframe using formula: x = z*standard deviation + mean:

for(i in names(Knn_NAs_preProcess$mean)){
  Knn_NAs_calculation[i] <- Knn_NAs_calculation[i]*Knn_NAs_preProcess$std[i] + Knn_NAs_preProcess$mean[i] 
}

# Let's calculate the accuracy of kNN:

na_index <- which(is.na(iris.mis[, -5]))
iris_vector <- unlist(iris[, -5]) # Given a list structure x, unlist simplifies it to produce a vector which contains all the atomic components which occur in x.
missing_values_original <- iris_vector[na_index]

iris_vector_imputed_knn <- unlist(Knn_NAs_calculation[, -5])
missing_values_imputed_knn <- iris_vector_imputed_knn[na_index]

# the root mean squared error between two numeric vectors
RMSE_kNN <- sqrt(mean(missing_values_imputed_knn - missing_values_original)^2)

```


2. NAs replaced by values calculated by randon forest:

```{r missing value calculation: randon forest with mice, echo=TRUE, message=FALSE, warning=FALSE}
# mice() generate the object of type mids
RF_NAs_calculation <-  mice(iris.mis, method="rf")  

# complete(): extracts the completed data from a 'mids' object
RF_NAs_calculation_result <- mice::complete(RF_NAs_calculation)

# Let's calculate the accuracy of randon forest:

iris_vector_imputed_rf <- unlist(RF_NAs_calculation_result[, -5])
missing_values_imputed_rf <- iris_vector_imputed_rf[na_index]

# the root mean squared error between two numeric vectors
RMSE_RF <- sqrt(mean(missing_values_imputed_rf - missing_values_original)^2)

```
