---
title: "Missing Value Analysis in R"
author: "Illarion Jabine"
date: "13/12/2019"
output:
  word_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
	echo = TRUE,
	message = FALSE,
	warning = FALSE
)
```


### Required packages:

* [missForest]: to generate NAs at random
* [mice]: for missing value patters
* [VIM]: Visualization and Imputation of Missing Values
* [dplyr]: select_if() - to select only numeric variables
* [Hmisc]: imputing missing values


## Missing Value Analysis and Handling

Missing values is a common problem, and might cause serious problem down the road during the model building phase.
So, handling this issue is of paramount importance. First of all you need to identify if a dataset contains any missing values. Once it's done you have 4 ways to handle them:
1. Go back to the source of data (if possible) and find out what causes the missing values.
2. Delete the observations with missing values.
3. Delete the variables with missing values.
4. Replace the missing values with the mean or median.
5. Predict missing values by imputing them using methods like kNN, random forest, tree, lm, etc.


As an example, let's create a dataset with missing values. We will use prodNA() function from lissForest package to generate random NAs:

```{r create dataframe with NAs}
library(missForest)
library(mice)
library(VIM)
library(dplyr)
library(tidyverse)
library(Hmisc)
data(iris)

# create a new dataframe with 10% missing values:
iris.mis <- prodNA(iris, .10)

# another way using sample() function to enject missing values:
iris.mis1 <- iris
iris.mis1[sample(1:nrow(iris), round(nrow(iris)/10)), "Sepal.Length"] <- NA
iris.mis1[sample(1:nrow(iris), round(nrow(iris)/10)), "Sepal.Width"] <- NA
iris.mis1[sample(1:nrow(iris), round(nrow(iris)/10)), "Petal.Length"] <- NA
iris.mis1[sample(1:nrow(iris), round(nrow(iris)/10)), "Petal.Width"] <- NA
```

## 1. Missing Value Initial Analysis

The first step is to analyse if the dataset contains missing values,what variables are affected and the distribution pattern.

The basic functions that allow to detect the presence of missing values incluce summary(), colSums()
```{r detecting missing values}

# shows the columns with missing values and their count.
summary(iris.mis)
# To see the total number of missing values:
sum(is.na(iris.mis))

# we select only numeric colums and checks if they contain NAs
select_if(iris.mis,is.numeric) %>% colSums()

# another way to select only numeric columns:
# iris.mis[, unlist(lapply(iris.mis, is.numeric))]

# More advanced way to see the structure of the missing data is to use md.pattern() from mice package:
mice::md.pattern(iris.mis)
# the last row of the matrix contains the number of missing values in each column

# To see missing values in more visual format use aggr() from VIM package:
# This function calculates or plots the amount of missing/imputed values in each variable and the amount of missing/imputed values in certain combinations of variables.

missing_value_analysis <- aggr(iris.mis, col=c("skyblue", "red"),
 numbers=TRUE, sortVars=TRUE,
 labels=names(iris.mis), cex.axis=.7,
 gap=2, ylab=c("Missing Data Histogram","Pattern"))

# To see summary of the missing values analysis produced by aggr use generic function summary():
summary(missing_value_analysis)

```

## 2.Replacing Missing Values with mean or median

Replacing NAs with mean or median values is quite a straightforward, brute force approach. It is not as sophisticated as predicting the missing values by means of prediction methodes, but as a quick and dirty approach it might do. 

```{r Replacing Missing Values with mean or median}
# Let's first exclude the categorical variable Species from our computation:
iris.mis <- iris.mis[,-5]

# Let's replace NAs with column vector means using mice() function from mice package:
mean_compute_mice <- mice::mice(data = iris.mis, method = "mean")

# To retrive the new dataset with computed NAs from the object mean_compute_mice use complete() function:
mean_compute_mice_res <- mice::complete(mean_compute_mice)

# We can impute NAs using impute() function from Hmisc package. If instead of function a single value is specified those value is used for insertion.

impute(iris.mis$Sepal.Length, mean) # replace NAs with mean
impute(iris.mis$Sepal.Length, median) # replace NAs with median
impute(iris.mis$Sepal.Length, 10) # replace NAs with 10

# Or we can replace missing values with minimal or maximal values:
impute(iris.mis$Sepal.Length, min)
impute(iris.mis$Sepal.Length, max)


# Replacing using R standard functionality:

# iris.mis$Sepal.Length[is.na(iris.mis$Sepal.Length)] <- mean(iris.mis$Sepal.Length, na.rm = TRUE)
```

## 3. Impute Missing Values by using prediction algorithm

This is more sophisticated methode of estimating values for the missing values by applying different prediction models learned for missing values.

For available imputation methodes, see help(mice).



